A Full Accounting of Landcover Map Error and Bias and Their Impacts on Assessments of Global Change

### 2015-10-01 paper/app_s1.tex
A number of methods revised and redone. Mostly described in new changelog file.

\begin{figure}[ht]
  \centering
     \includegraphics[width = 16cm]{figures/biases_1-100km.pdf} 
%     \vspace{-3cm}
      \caption{Biases and mean absolute errors (MAE) for each of the cropland maps as a function of cropland density (calculated using the 2011 reference maps) and aggregation scales. Rows present biases by map product, columns by aggregation scale.  Dash lines indicate bias at each level of cropland density, calculated in bins spanning 5\% of density (e.g. 0-5\% cropland cover, 5-10\%, etc.), while solid lines indicate the mean absolute error.  The black numbers in each plot area present the overall means of bias/MAE for each sensor-scale combination. }
      \label{fig:default}
\end{figure}

@@

\begin{figure}[ht]
  \centering
     \includegraphics[width = 12cm]{figures/biases_1km.pdf} 
%     \vspace{-3cm}
      \caption{Biases and mean absolute errors (MAE) for each of the cropland maps at 1 km resolution, as a function of cropland density. Colored lines (color-coded to map product name) show the bias/MAE at each level of cropland density, calculated in bins spanning 5\% (e.g. 0-5\% cropland cover, 5-10\%, etc.). Box plots show the variability of bias in each bin (whiskers = 2.5 and 97.5 percentiles, box the inter-quartile, and grey bar in box the median). Biases are presented in the top row, and MAEs in the bottom row. Statistics are calculated from pooled map errors calculated from differences between the 2007 reference map and each cropland map, and the 2011 reference map and each cropland map.}
      \label{fig:default}
\end{figure}

@@

\begin{figure}[ht]
  \centering
     \includegraphics[width = 12cm]{figures/yld_bias_map_2.pdf} 
      \caption{Errors (normalized to the reference-derived country mean) in disaggregated maize yield estimates.}
      \label{fig:default}
\end{figure}

\begin{figure}[ht]
  \centering
     \includegraphics[width = 12cm]{figures/prod_bias_map_2.pdf} 
      \caption{Errors (normalized to reference-derived country mean) production estimates calculated from disaggregated maize yield and harvested area estimates.}
      \label{fig:default}
\end{figure}

>>>>

\input{figures/cropland2011-bias.tex}

\begin{figure}[ht]
  \centering
     \includegraphics[width = 16cm]{figures/cropland_bins.pdf} 
%     \vspace{-3cm}
      \caption{Number of cells within each cropland density bin at each scale of aggregation, where bins represent 5\% increment of cropland cover from 0 to 100\% (values on x-axis provide the upper limit of each bin).}
      \label{fig:default}
\end{figure}

\input{figures/cropland2007-bias.tex}

\begin{figure}[ht]
  \centering
     \includegraphics[width = 16cm]{figures/biases_1-100km.pdf} 
%     \vspace{-3cm}
      \caption{Biases and mean absolute errors (MAE) for each of the cropland maps as a function of cropland density (calculated using the 2011 reference maps) and aggregation scales. Rows present biases by map product, columns by aggregation scale.  Dash lines indicate bias at each level of cropland density, calculated in bins spanning 5\% of density (e.g. 0-5\% cropland cover, 5-10\%, etc.), while solid lines indicate the mean absolute error.  The black numbers in each plot area present the overall means of bias/MAE for each sensor-scale combination. The bin-wise and overall mean statistics were calculated from pooled map errors calculated from differences between the 2007 reference map and each cropland map (including all three variants--high, medium, and low--of the MODIS and GlobCover-derived cropland maps), and the 2011 reference map and each cropland map. }
      \label{fig:default}
\end{figure}

@@

\begin{figure}[ht]
  \centering
     \includegraphics[width = 12cm]{figures/biases_1km.pdf} 
%     \vspace{-3cm}
      \caption{Biases and mean absolute errors (MAE) for each of the cropland maps at 1 km resolution, as a function of cropland density. Colored lines (color-coded to map product name) show the bias/MAE at each level of cropland density, calculated in bins spanning 5\% (e.g. 0-5\% cropland cover, 5-10\%, etc.). Box plots show the variability of bias in each bin (whiskers = 2.5 and 97.5 percentiles, box the inter-quartile, and grey bar in box the median). Biases are presented in the top row, and MAEs in the bottom row. Statistics were calculated from pooled map errors calculated from differences between the 2007 reference map and each cropland map (including all three variants--high, medium, and low--of the MODIS and GlobCover-derived cropland maps), and the 2011 reference map and each cropland map. }
      \label{fig:default}
\end{figure}

@@

\FloatBarrier
\input{figures/cropadj-mae.tex}

\clearpage
\begin{figure}[ht]
  \centering
     \includegraphics[width = 12cm]{figures/yld_bias_map_1.pdf} 
      \caption{Errors (normalized to the reference-derived country mean) in disaggregated maize yield estimates.}
      \label{fig:default}
\end{figure}

\begin{figure}[ht]
  \centering
     \includegraphics[width = 12cm]{figures/prod_bias_map_1.pdf} 
      \caption{Errors (normalized to reference-derived country mean) production estimates calculated from disaggregated maize yield and harvested area estimates.}
      \label{fig:default}
\end{figure}

### 2015-10-01 paper/main.tex
A number of methods revised and redone. Mostly described in new changelog file.

\dropcap{T}he nature and distribution of landcover is an indicator that gives significant insight into socio-econonomic processes \cite{lambin_modelling_1997}, because human endeavors are so closely tied to how we transform land, whether it be the felling of ancient forests for farmland or erecting skycrapers. The vastness of our alteration of Earth’s landscapes suggests that landcover is a fundamental mediator of many environmental and social processes that drive or are affected by global change \cite{lambin_modelling_1997}, such as agricultural production and food security \cite{lark_cropland_2015,wright_recent_2013, licker_mind_2010}, carbon cycling \cite{asner_high-resolution_2010,gaveau_major_2014}, biodiversity loss \cite{newbold_global_2015,luoto_predicting_2004}, and changes in human demography \cite{linard_assessing_2010}. Like any view into nature, resolution and fidelity at fine scales are the keys to unlocking more granular and mechanistic insights into these processes \cite{see_improved_2015}. It is therefore unsurprising to see the explosive growth in private sector initiatives to develop new Earth observing capabilities, which range from small hobbyist drones\footnote{e.g. 3DRobotics, DJIA} to satellite arrays\footnote{Planet Labs, Skybox}, in order to add value to industries such as agriculture, mining, and construction. This rapid growth in fine-scale landcover mapping capability is creating new opportunities to develop actionable information for traditionally public-sector concerns, such as agricultural development\footnote{USAID's Feed the Future}, drought and flood adaptation\footnote{Global Index Insurance Facility, www.indexinsuranceforum.org}, and carbon cycle management\footnote{United Nations REDD+, www.un-redd.org/aboutredd}. But while the demand for more nuanced, landcover-based insights is growing, there is only now the opportunity to use finer-scaled imagery to comprehensively interrogate the accuracy and biases in the landcover products that have become ubiquitous in global change research.

Global landcover data can only practically be derived from satellite imaging, but in many regions the average size class of the cover type of interest is smaller than the sensor resolution, or spectrally indistinct from other neighboring covers, which propagates classification error \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. The result is that landcover maps are generally inaccurate at finer scales and disagree greatly, particularly in those parts of the world undergoing the most rapid land use changes, where these sources of bias tend to be most pronounced \cite{estes_projected_2013,fritz_comparison_2010,fritz_cropland_2011}.

@@

Just as a tall edifice cannot be built on a shaky foundation, global change science needs to be based on sound landcover data. There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need in this study, using a unique, high accuracy agricultural landcover map for the entire country of South Africa to conduct a spatially comprehensive, bottom-up quantification of error in several latest generation landcover maps that are widely used in global change studies. We use these errors to assess the extent of bias in i) landcover data, ii) how  landcover properties influence this bias, iii) how these biases change with aggregation scale, with the specific goal of determining ``safe'' scales for drawing area-based inferences, and iv) how these biases propagate through several different forms of downstream analyses that broadly represent the global change research focus areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

@@

\vspace{-0.5 cm}
\section{Quantifying Error and Bias}
We first quantified the errors in cropland area estimates based on the pixel-wise differences between the reference map and each of the four cropland maps. We calculated these errors for five different scales of aggregation, from the original 1 km up to 100 km, in order to calculate bias (the mean pixel-wise error) and mean absolute error (which measures the average error magnitude), and how these vary with scale. Next, we assessed the degree to which the average cropland cover in agricultural landscapes, which relates to their pattern, impacts error magnitude.

@@

We created the 1 km reference map after removing all field types classified as communal/smallholder agriculture (individual fields in this category were not mapped, thus they were removed to prevent potential cropland area overestimates) or permanent tree-crops (SI), and calculated the total cropland extent in the remaining area (1,081,000 km$^2$, or 90\% of South Africa). The 2011 reference map showed a cropland area of 104,310 km$^2$, which the SA-LC and GeoWiki maps overestimated by 26 and 5.8\%, respectively, and GlobCover and MODIS underestimated by 21 and 26.1\%.

We then aggregated the reference and each cropland map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from the reference map at each scale of aggregation to assess error patterns (Fig. 1). Negative pixels here represent overestimation error, while positive values indicate underestimates.

@@

The most pronounced errors were in the MODIS and GlobCover maps, which both underestimated cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimated along the eastern to northern margins (red areas in Fig. 1). These average of these were 21\% for MODIS and 34\% for GlobCover at 1 km resolution (Fig. S1), meaning that each map had a strong underestimation bias. Both maps's biases decreased with pixel aggregation, with MODIS biases falling to 8\% at 50 km, but GlobCover's was still 24\% at 100 km (SI Appendix, Fig. S1).

@@

\section{Discussion}

@@

1. Previous studies have quantified disagreement over large areas, where

>>>>

\dropcap{T}he nature and distribution of landcover provides significant insight into econonomic processes \cite{lambin_modelling_1997} because human endeavors are so closely tied to how we transform land, whether it be the felling of ancient forests for farmland or converting that farmland into office parks. The vastness of our alteration of Earth’s landscapes suggests that landcover is a prime mediator of many environmental and social processes that drive or are affected by global change \cite{lambin_modelling_1997}, such as agricultural production and food security \cite{lark_cropland_2015,wright_recent_2013, licker_mind_2010}, carbon cycling \cite{asner_high-resolution_2010,gaveau_major_2014}, biodiversity loss \cite{newbold_global_2015,luoto_predicting_2004}, and changes in human demography \cite{linard_assessing_2010}. Like any view into nature, resolution and fidelity at fine scales are the keys to unlocking more granular and mechanistic insights into these processes \cite{see_improved_2015}. It is therefore unsurprising to see the explosive growth in private sector initiatives to develop new Earth observing capabilities, which range from small hobbyist drones\footnote{e.g. 3DRobotics, DJIA} to satellite arrays\footnote{Planet Labs, Skybox}, in order to add value to industries such as agriculture, mining, and construction. This rapid growth in fine-scale landcover mapping capability is creating new opportunities to develop actionable information for traditionally public-sector concerns, such as agricultural development\footnote{USAID's Feed the Future}, drought and flood adaptation\footnote{Global Index Insurance Facility, www.indexinsuranceforum.org}, and carbon cycle management\footnote{United Nations REDD+, www.un-redd.org/aboutredd}. But while the demand for more nuanced, landcover-based insights is growing, there is only now the opportunity to use finer-scaled imagery to comprehensively interrogate the accuracy and biases in the landcover products that have become ubiquitous in global change research.

Global landcover data can only practically be derived from satellite imaging, but in many regions the cover types of interest are smaller, on average, than the sensor resolution, or spectrally indistinct from other neighboring covers, and these factors propagate classification errors \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. The result is that landcover maps are generally inaccurate at finer scales and disagree substantially with one another, particularly in those parts of the world undergoing the most rapid land use changes \cite{estes_projected_2013,fritz_comparison_2010,fritz_cropland_2011}.

@@

Just as a building needs a solid foundation, global change science needs to be based on sound landcover data. There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need in this study, using a unique, high accuracy agricultural landcover map for the entire country of South Africa to conduct a spatially comprehensive, bottom-up quantification of error in several latest generation landcover maps that are widely used in global change studies. We use these errors to assess the extent of bias in i) landcover data, ii) how  landcover properties influence this bias, iii) how these biases change with aggregation scale, with the specific goal of determining ``safe'' scales for drawing area-based inferences, and iv) how these biases propagate through several different forms of downstream analyses that broadly represent the global change research focus areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

@@

\vspace{-0.5 cm}
\section{Quantifying Error and Bias}
We first quantified the errors in cropland area estimates based on the pixel-wise differences between the reference map and each of the four cropland maps. We calculated these errors for five different scales of aggregation, from the original 1 km up to 100 km, in order to calculate bias (the mean pixel-wise error) and mean absolute error (which measures the average error magnitude), and how these vary with scale. Next, we assessed the degree to which the average cropland cover in agricultural landscapes, a descriptor of landcover pattern, impacts error magnitude.

@@

We created the 1 km reference map after removing all field types classified as communal/smallholder agriculture (individual fields in this category were not delineated, thus they were removed to prevent potential cropland area overestimates, SI) or permanent tree-crops (SI), and calculated the total cropland extent in the remaining area (1,081,000 km$^2$, or 90\% of South Africa). The 2011 reference map showed a cropland area of 104,310 km$^2$, which the SA-LC and GeoWiki maps overestimated by 26 and 5.8\%, respectively, and GlobCover and MODIS underestimated by 21 and 26.1\%.

We then aggregated the reference and each cropland map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from each reference (2007 and 2011) map at each scale of aggregation to assess error patterns (Fig. 1). Negative pixels here represent overestimation error, while positive values indicate underestimates.

@@

The most pronounced errors were in the MODIS and GlobCover maps, which both underestimated cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimated along the eastern to northern margins (red areas in Fig. 1). Averaging these errors across the whole country shows resulted

The average of these were 21\% for MODIS and 34\% for GlobCover at 1 km resolution (Fig. S1), meaning that each map had a strong underestimation bias. Both maps's biases decreased with pixel aggregation, with MODIS biases falling to 8\% at 50 km, whereas GlobCover's was still 24\% at 100 km (SI Appendix, Fig. S1).

@@

\section{Discussion}
This spatially comprehensive, bottom-up assessment of error and biases in landcover maps provides unique insight into their extent, causes, and consequences for understanding global change processes, made possible by a unique, high accuracy dataset that provides the truest measure of total cropland area and distribution that is currently available for this region. This dataset is of course not perfect, being affected by the map-makers' occasional interpretation errors (mostly of omission), while some of the cropland map error we found may have been caused by the slight temporal mismatches between the reference data and the original landcover datasets we used. However, our assessment (SI) suggests that these errors are small, and do not have an appreciable impact on our general findings about the degree of error and bias in landcover data, which is bolstered by previous work showing large disagreements between landcover map-based cropland area estimates and national inventory data \cite{fritz_comparison_2010}.

These results suggest several guidelines for selecting and using landcover data, and contain some important implications for how understanding of global change processes based on the data, and associated policy decisions, may be affected. In terms of developing a base landcover map, the first important rule of thumb is that standard landcover products derived from coarse resolution sensors, such as MODIS and GlobCover, appear to be too biased to be useful without substantial aggregation. If we use the standard that bias within +/-10\% is acceptable, then at least 25-100 km of aggregation is needed to sufficiently cancel out the errors in the base landcover data and subsequent first order estimates built on them (Fig. 3 \& 4). The upper range of aggregation scale is necessary if a mixed pixel class becomes dominant, as in the case with GlobCover, because these lead to underestimation bias that will persist until the pixel size becomes substantially greater than the average area of landscapes that are dominated by the cover type of interest, which can be $>$1000 km$^2$ in some of South Africa's farming regions.

Maps derived from higher resolution sensors, such as the SA-LC dataset, if carefully done, do not have this mixed class problem, and  sufficiently unbiased for most applications with just 1-5 km of aggregation. However, such datasets are typically developed for specific countries, using varying methods, and can be hard to obtain. For broader scale analyses, the best option is to use newer generation maps such as GeoWiki (and the GLC-Share \footnote{GLC-Share. www.glcn.org} datasets for other cover types) which is relatively unbiased at 1 km resolution. GeoWiki's lower bias comes from its process of evaluating consensus between several landcover datasets (including the other three in this study), resulting in cropland probabilities that are converted to percentages by calibrating to statistical data \cite{fritz_cropland_2011,fritz_mapping_2015}. This method mirrors the ensemble averaging used by other fields (e.g. crop \cite{asseng_uncertainty_2013}, climate \cite{giorgi_calculation_2002}, and ecological modeling \cite{araujo_ensemble_2007}) to increase prediction confidence.

The statistical constraint procedure is similar to the one we used \cite[following][]{ramankutty_farming_2008}, which resulted in maize unbiased production estimates (Fig. 3) because it eliminated bias in the adjusted cropland and harvested maps that they were built upon. This result, together with GeoWiki's low bias, indicate the value of fusing inventory data with remote sensing. However, this approach depends on the quality of inventory data, which can be suspect in many countries, particularly in Africa \cite{carletto_emperor_2013,fao_action_2013}. The statistical constraint also does not greatly improve map \emph{accuracy}, as evidenced by GeoWiki's 23\% mean absolute error in 1 km cropland percentage estimates (Fig. S1), which is only slightly more accurate than MODIS (31\%) but worse than SA-LC (11\%). GeoWiki is definitely most accurate among the large scale landcover products, but its improvement is related to the map consensus methods, which can correct for omission or commission errors made by the classifier. Statistical constraints only adjust map values at locations where cropland is identified, so their use it .

Map accuracy is perhaps more important than bias for landcover maps.

Broader regional implications - error higher elsewhere

@@
